---
title: "CookBook.Rmd"
output: html_document
---

This code book describes a project focused upon producing a tidy data set that describes the mean by activity and subject of all the mean and standard deviation observations recorded in the data sets provided at: https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip. In this CookBook book we describes the variables, the data, and the work that was performed to clean up the data and ultimately produce our tidy dataset.

# Source Data

Using the dataset located at :https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip I have read the train and test datasets. The code snippet below shows the parts of the source data used for this exercise.

```{r}

  #
  # read the test data in and build a consolidated table.
  #
  test_data     <-  read.table('UCI HAR Dataset/test/X_test.txt', header=FALSE)
  test_subject  <-  read.table('UCI HAR Dataset/test/subject_test.txt', header=FALSE)
  test_activity <-  read.table('UCI HAR Dataset/test/y_test.txt', header=FALSE)
  #
  # Column bind the parts into a single test dataset
  #
  all_test <- cbind( test_subject, test_activity, test_data )
  #
  # now pull together the training data..
  #
  train_data     <- read.table('UCI HAR Dataset/train/X_train.txt', header=FALSE)
  train_subject  <- read.table('UCI HAR Dataset/train/subject_train.txt', header=FALSE)
  train_activity <- read.table('UCI HAR Dataset/train/y_train.txt', header=FALSE)
  #
  # Column bind all the training data into a single dataset
  #
  all_train <- cbind( train_subject, train_activity, train_data)

```

At the end of this activity, we have two complete subsets of our ultimate dataset that we have produced using the `read.csv()` and `cbind()` functions.

## Complete Data set

I combined them using `cbind()` and `rbind()` as follows in order to have a complete data setup upon which to process...

```{r}
  #
  # Column bind all the training data into a single dataset
  #
  all_train <- cbind( train_subject, train_activity, train_data)
  #
  # unify the test and training data using row binding...
  #
  complete_data_set <- rbind( all_test, all_train )


```

# Variables

The table below shows the data that was identified as being either a mean or a standard deviation measurement. These I pulled from the source data (complete_data_set) and created a subset of the data.

Column | Variable name
------  ---------------------
1  | tBodyAcc-mean()-X
2 | tBodyAcc-mean()-Y
3 | tBodyAcc-mean()-Z
4 | tBodyAcc-std()-X
5 | tBodyAcc-std()-Y
6 | tBodyAcc-std()-Z
41 | tGravityAcc-mean()-X
42 | tGravityAcc-mean()-Y
43 | tGravityAcc-mean()-Z
44 | tGravityAcc-std()-X
45 | tGravityAcc-std()-Y
46 | tGravityAcc-std()-Z
81 | tBodyAccJerk-mean()-X
82 | tBodyAccJerk-mean()-Y
83 | tBodyAccJerk-mean()-Z
84 | tBodyAccJerk-std()-X
85 | tBodyAccJerk-std()-Y
86 | tBodyAccJerk-std()-Z
121 | tBodyGyro-mean()-X
122 | tBodyGyro-mean()-Y
123 | tBodyGyro-mean()-Z
124 | tBodyGyro-std()-X
125 | tBodyGyro-std()-Y
126 | tBodyGyro-std()-Z
161 | tBodyGyroJerk-mean()-X
162 | tBodyGyroJerk-mean()-Y
163 | tBodyGyroJerk-mean()-Z
164 | tBodyGyroJerk-std()-X
165 | tBodyGyroJerk-std()-Y
166 | tBodyGyroJerk-std()-Z
201 | tBodyAccMag-mean()
202 | tBodyAccMag-std()
214 | tGravityAccMag-mean()
215 | tGravityAccMag-std()
227 | tBodyAccJerkMag-mean()
228 | tBodyAccJerkMag-std()
240 | tBodyGyroMag-mean()
241 | tBodyGyroMag-std()
253 | tBodyGyroJerkMag-mean()
254 | tBodyGyroJerkMag-std()
266 | fBodyAcc-mean()-X
267 | fBodyAcc-mean()-Y
268 | fBodyAcc-mean()-Z
269 | fBodyAcc-std()-X
270 | fBodyAcc-std()-Y
271 | fBodyAcc-std()-Z
294 | fBodyAcc-meanFreq()-X
295 | fBodyAcc-meanFreq()-Y
296 | fBodyAcc-meanFreq()-Z
345 | fBodyAccJerk-mean()-X
346 | fBodyAccJerk-mean()-Y
347 | fBodyAccJerk-mean()-Z
348 | fBodyAccJerk-std()-X
349 | fBodyAccJerk-std()-Y
350 | fBodyAccJerk-std()-Z
373 | fBodyAccJerk-meanFreq()-X
374 | fBodyAccJerk-meanFreq()-Y
375 | fBodyAccJerk-meanFreq()-Z
424 | fBodyGyro-mean()-X
425 | fBodyGyro-mean()-Y
426 | fBodyGyro-mean()-Z
427 | fBodyGyro-std()-X
428 | fBodyGyro-std()-Y
429 | fBodyGyro-std()-Z
452 | fBodyGyro-meanFreq()-X
453 | fBodyGyro-meanFreq()-Y
454 | fBodyGyro-meanFreq()-Z
503 | fBodyAccMag-mean()
504 | fBodyAccMag-std()
513 | fBodyAccMag-meanFreq()
516 | fBodyBodyAccJerkMag-mean()
517 | fBodyBodyAccJerkMag-std()
529 | fBodyBodyGyroMag-mean()
530 | fBodyBodyGyroMag-std()
539 | fBodyBodyGyroMag-meanFreq()
542 | fBodyBodyGyroJerkMag-mean()
543 | fBodyBodyGyroJerkMag-std()
552 | fBodyBodyGyroJerkMag-meanFreq()

# Transformations

The code snippett below demonstrates how the subset of the source data was created -- note the offsets are pushed by (2 columns) because of the addition of the subject and activity columns we I created the original complete data set.

```{r}
 #
 # using offsets in meansandstd measures.txt (offset by 2 cols since we added a subject
 # activities column to the dataset) we setup our subset of the dataset.
 #
  mean_and_std_data <- complete_data_set[c(1:2, 3:8, 43:48, 83:88, 
                                           123:128, 163:168, 203:204,
                                           216:217, 229:230, 242:243, 
                                           255:256, 268:273, 
                                           296:298, 347:352, 
                                           375:377, 426:431, 
                                           454:456, 505:506, 515, 518:519, 531:532,
                                           541, 544:545, 555)]  
```

In order to produce the final tidy dataset - I named all the columns in the mean_and_std_data set to the name reflected in the table above. Then performed two transformations on the dataset... I melted the data in order to identify the id variables as subject and activity, then cast the dataset by subject and means per activty -- this was done using the code below:

```{r}
##
 # Melt our dataset, such that the variables are all flattened into one column
 #
 new_data <- melt( mean_and_std_data, id.vars= c("Subject", "Activity"))
 ##
 # get the mean by variable, and subject and activity.
 #
 meansOfSubjectByActivtity <- dcast(new_data, Subject + Activity ~ variable, mean)
```

# Tidy Dataset creation

Finally I wrote the dataset to a file named: MeansBySubjectAndActivity.txt which the script will put in your current working directory at the successful completion of it. The snippet below demonstrates the write.file() used to produce the final tidy dataset.

```{r}

##
 ## write this table out as a tidy dataset.
 ##
 write.table(meansOfSubjectByActivtity, "MeansBySubjectAndActivity.txt", 
              row.name=FALSE, 
             sep="\t")
```
